---
title: "First R Exercises"
output: html_document
---

Show all code and outputs in any problems below that use R. 3 points per problem. Show all of your work. Read each question completely before answering. You'd be surprised how many people miss details...

0. Free 3 points - complete this homework in Rmarkdown!  

1. Create a vector of 100 randomly distributed numbers between 0 and 100 using `runif` and save the vector into the variable `my_vec`. If you cannot figure out runif, try `?runif`. What information does `str` and `summary` tell you about `my_vec`?  How do they differ?  

```{r echo=FALSE, eval=FALSE}
my_vec <- runif(100, 0, 100)

str(my_vec)
summary(my_vec)

#str shows what information is in the vector and its length. summary 
#gives information about the distribution of the data
```

2. Try out a little plot. What does `hist(my_vec)` show you? What information does the helpfile for `hist` tell you about what you just did?  
```{r echo=FALSE, eval=FALSE}
hist(my_vec)

#This is a histogram of my_vec
#showing frequently data occurs in a set of bins
```

3. Load up the `mtcars` data set using `data(mtcars)`.  Apply the following functions to mtcars: `class`, `str`, `summary`.  What do these three functions tell you about mtcars?  

```{r echo=FALSE, eval=FALSE}
data(mtcars)
class(mtcars)
str(mtcars)
summary(mtcars)


#mtcars is a data frame with 32 values in 11 columns. The values
#are numeric, and summary shows me how they are distributed
```

4. Look at the help file for the class `data.frame`.  What does it tell you about these objects?  

```{r echo=FALSE, eval=FALSE}
#data.frames are both lists and matrices, and are fundamental to most of the modeling
#in R
```

5. What  are you thinking about working with for your final project? Give me a brief description of your current idea. I know this might be early, but I'm guessing you've got something in mind.