---  
  title: "Course Schedule and Readings"  
---  

<!--


NEXT TIME I TEACH 607 in 2024
redo explaining least squares - it's not good in the lecture

re-arrange the linear regression lab
  - workflow
  - transforms
  - viz of prediction intervals and such
-->

  While the topics covered are broad, each week will feature different examples from genetics, ecology, molecular, and evolutionary biology highlighting uses of each individual set of techniques.    
  
  F = Fieberg, R4DS = [Wickham et al.](https://r4ds.hadley.nz/), U/P for linked pdfs = biol607  
  
  Change .html to .Rmd in your browser to get the markdown of all lectures, etc.

**Recorded Lecture**: All lectures and labs will be recorded and are available at https://www.youtube.com/playlist?list=PLZRMqMK8aRmIMuiEX-QVxLNk2o6SlRa-m   
    
**Turning in Homework**: All homework should be completed using RMarkdown or Quarto. You'll freely mix answers in text and code there. Use projects. Please zip up the full project folder for each to submit so that we have the .rmd (or .qmd), .html output, data, etc. and can recompile your homework if we need to. An ideal project structure would be something like this:
  
homework_X/  
&nbsp; |--- markdown  
&nbsp; |--- scripts  
&nbsp; |--- data  

directory structure, so that all data is in `data/`. **Please zip up the archive for the homework and standardize filenames as follows: `number_lastName_firstName_2023.zip`** where number is the week number the homework is due (from weeks below) and make sure to include the 0s for numbers like 01), and your last and first names - well, you should know them!  
  
To submit homework, use the dropbox link you were sent by the TA.  


```{r dates, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
current_date = parse_date_time("2023-09-04", orders="ymd")
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```    
     
## Block 1: Introduction to Programming and Reproducibility

### Week 1. 
`r datestring`   
__Lecture:__ [Class intro](lectures/01_intro_x.html), [Intro to R](lectures/02_r_intro.html). \
__Lab:__ [Matrices, Lists and Data Frames](lectures/02a_r_intro_list_matrix_df.html). [Introduction to Markdown](https://rmarkdown.rstudio.com/lesson-1.html)    
__Reading:__ R4DS [Intro](https://r4ds.hadley.nz/intro), [Workflow basics](https://r4ds.hadley.nz/workflow-basics), [Scripts and Projects](https://r4ds.hadley.nz/workflow-scripts), [Code Style](https://r4ds.hadley.nz/workflow-style), [Vectors](http://r4ds.had.co.nz/vectors.html), and [Quarto](https://r4ds.hadley.nz/quarto)   
__Further Exploration:__ [Quarto Formats](https://r4ds.hadley.nz/quarto-formats), https://quarto.org/ for everything!  
__Cheat Sheets:__ [Quarto Cheat Sheet](https://rstudio.github.io/cheatsheets/quarto.pdf)  
__Install R:__ Go to https://cloud.r-project.org/ and get the right version of R for you. Then, go to https://www.rstudio.com/products/rstudio/download/#download and install Rstudio.  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-intro-2023  
__Homework:__ [Intro to R and Data Frames](./homework/intro_to_r.html)  

### Week 2.   
```{r next_date, echo=FALSE}
current_date <- current_date+ days(7)
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```
`r datestring`   
__Lecture:__  [Data visualization](lectures/05_data_viz_principles.pptx)  
__Lab Topic:__ [Introduction to ggplot2](lab/04_ggplot_palmer_intro.html).  
__Reading:__ R4DS Chapters on [Data Vizualization](https://r4ds.hadley.nz/data-visualize), [Layers of Plots](https://r4ds.hadley.nz/layers), and [Graphics for Communication](https://r4ds.hadley.nz/communication),  [Introducing Palmer Penguins](https://journal.r-project.org/articles/RJ-2022-020/)    
__Further Exploration:__  [Friendly 2008 on History of Data Viz](http://byrneslab.net/classes/biol607/readings/Friendly_2008_dataviz_history.pdf), [Unwin 2008](http://byrneslab.net/classes/biol607/readings/Unwin_2008_dataviz.pdf), [DC Starting with Data](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html), [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/) - note, this is a whole book, but scan it - it's amazing\
__Etherpad:__ https://etherpad.wikimedia.org/p/dataviz-2023  
__Packages used this Week:__ [ggplot2](https://ggplot2.tidyverse.org/), [ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html), [ggdist](https://mjskay.github.io/ggdist/) - `install.packages(c("ggplot2", "ggridges", "palmerpenguins", "ggdist"))`  
__Cheat Sheets:__  [Ggplot2 cheat sheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf), [Choosing a good chart cheat sheet](https://biol355.github.io/Readings/choosing-a-good-chart-09.pdf)  
__Homework:__ [ggplot2 homework](./homework/ggplot.html)  
  
### Week 3.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture/Lab:__  [Data import, using libraries](https://datacarpentry.org/R-ecology-lesson/02-starting-with-data.html),  [Pipes, data manipulation, summarization, and making it tidy](https://datacarpentry.org/R-ecology-lesson/03-dplyr.html)    
<!--, [Joins](lectures/18_join.html)  -->
__Data:__ [Portal Data](https://ndownloader.figshare.com/files/2292169) - and learn more [here](https://portal.weecology.org/)  
__Reading:__  [Data organization in spreadsheets](./readings/Browman_and_Woo_Spreadsheets.pdf),  R4DS Chapters on [data import](https://r4ds.hadley.nz/spreadsheets),  [pipes](https://r4ds.hadley.nz/workflow-style.html#sec-pipes), [data transformation](https://r4ds.hadley.nz/data-transform.html), [tidy data](https://r4ds.hadley.nz/data-tidy.html)    
__Optional Reading:__ [10 Commandments for Good Data Managament](https://dynamicecology.wordpress.com/2016/08/22/ten-commandments-for-good-data-management/), [Managing Data Frames with the Dplyr package](https://bookdown.org/rdpeng/exdata/managing-data-frames-with-the-dplyr-package.html), [Strings](http://r4ds.had.co.nz/strings.html), [factors](https://r4ds.hadley.nz/factors.html), and [Dates](http://r4ds.had.co.nz/dates-and-times.html)   
__Cheat Sheets:__ [Reading data into R](https://rstudio.github.io/cheatsheets/data-import.pdf), [Dplyr cheat sheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf).  
__Packages:__ `install.packages(c("dplyr", "janitor", "skimr", "lubridate", "tidyr", "readr", "readxl", "tibble"))` - [readr](https://readr.tidyverse.org/),  [readxl](https://readxl.tidyverse.org/),  [tibble](https://tibble.tidyverse.org/),  [skimr](https://docs.ropensci.org/skimr/), 
[janitor](https://garthtarr.github.io/meatR/janitor.html),
[visdat](https://docs.ropensci.org/visdat/)  
__Etherpad:__ http://etherpad.wikimedia.org/p/607-tidy-2023  
__Homework:__ [Data Manipulation and Pivot homework](./homework/dplyr_tidyr.html)  


### Week 4.   
```{r next_date, echo=FALSE }
```
`r datestring`   
__Lecture:__  [What is a Sample?](lectures/03_sampling_lecture_x.html) and [Sampling Distribution](lectures/04_sampling_dist.html).   
__Lab Topic:__ [Sampling and simulation](lab/sample_distributions_simulation.html).    
__Reading:__ [Cumming et al. 2007 on SDs, SEs, and CIs](http://byrneslab.net/classes/biol607/readings/Cumming_2007_error.pdf)   
__Etherpad:__ https://etherpad.wikimedia.org/p/sampling-2023  
__Packages for the Week:__ [dplyr](https://dplyr.tidyverse.org/) - `install.packages(c("dplyr", "purrr"))`  
__Homework:__ [Sampling, Simulation, and Tidy Tuesday](./homework/sampling_dist_tidytuesday.html)  


## Block 2: Linear Models for Data Analysis

### Week 5.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Introduction to Regression: [Correlation and Regression](lectures/linear_regression.html), [Fit and Precision](lectures/linear_regression_details.html)  
__Lab Topic:__ [Linear regression, diagnostics, visualization](lab/06_lm.html), and [data](lab/data_06.zip) \
__Reading:__ [Fieberg linear regression review](https://fw8051statistics4ecologists.netlify.app/linreg.html), R4DS on [model basics](http://r4ds.had.co.nz/model-basics.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-lm-2023  
<!--__In Class Code:__ [reproducible examples](in_class_code/2020/scripts/reprex.R), [lm](in_class_code/2020/scripts/linear_regression.R)  -->
__Homework:__ [Correlation and Linear Models](homework/06_correlation_regression.html)  


### Week 6   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__  [Many Predictors](./lectures/mlr.html),  [Categorical Predictors: Just Another Linear Model](lectures/categorical_predictors.html)   
__Lab Topic:__ [Models with Categorical Variables](lab/multiple_predictors.html)  
__Lab Data:__  [Multiple Files](lab/data/categorical_data.zip)   
__Reading:__ [Feiberg 3](https://fw8051statistics4ecologists.netlify.app/matrixreg), [Common statistical tests are linear models](https://lindeloev.github.io/tests-as-linear/), [Feiberg ch 6](https://fw8051statistics4ecologists.netlify.app/multicol.html)   
__Optional Reading:__ [Analysis of variance with unbalanced data: an update for ecology & evolution](./readings/Hector_et_al-2010-Journal_of_Animal_Ecology.pdf), [Day and Quinn 1989 on Post-hocs](https://www.jstor.org/stable/1943075).   
__Packages for The Week:__ `install.packages(c("car", "emmeans", "multcompView", "contrast", "visreg"))`    
__Etherpad:__ https://etherpad.wikimedia.org/p/607-many-predictors-2023  
__Homework:__ None. Catch up!    

<!-- __In Class Code:__ [comparing two means](./in_class_code/2020/scripts/comparing_means.R)  -->  


### Week 7   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lectures:__  [Many categories or mixing categories and continuous predictors](./lectures/many_types_of_predictors.html), [When one predictor's effect depends on the other](lectures/interactions_lm.html)   
__Lab Topic:__  [Models with many types of predictors - and interactions!](lab/complex_linear_models.html)  
__Lab Data:__  [Multiple Files](lab/data_10.zip)  
__Reading:__  [Feiberg ch 6](https://fw8051statistics4ecologists.netlify.app/multicol.html),  [Simple means to improve the interpretability of regression coefficients](./readings/Schielzeth_2010_MEE.pdf), [Understanding ‘it depends’ in ecology: a guide to hypothesising, visualising and interpreting statistical interactions](https://onlinelibrary.wiley.com/doi/full/10.1111/brv.12939)   
__Optional Readings:__ [Interactions in statistical models: Three things to know](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13714), [Centring in regression analyses: a strategy to prevent errors in statistical inference](./readings/kramer_blasey_centering.pdf)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-complex-linear-models-2023  
__Homework:__ [Complex linear models](homework/complex_linear_models.html)    
__Project Proposal:__ [general instructions here](project_proposal.html)  


### Week 8   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ [Generalized Linear Models](lectures/generalized_linear_models.html), [Fitting, Count Data,  and Overdispersion](lectures/glm_overdispersion.html)     
__Data:__ [GLS Data](./lectures/data/gls/gls_data.zip), [GLM Data](./lab/gzlm_data.zip)  
__Lab Topic:__ [Generalized Linear Models](./lab/12_gzlm.html)  
__Packages for the Week:__ `install.packages(c("MASS", "readxl", "betareg", "DHARMa", "TMB", "glmmTMB"))`  
__Reading:__ [Feiberg on Generalized Linear Models](https://fw8051statistics4ecologists.netlify.app/glm.html), [Models for Count Data](https://fw8051statistics4ecologists.netlify.app/poissonr), [Feiberg on Logistic Regression](https://fw8051statistics4ecologists.netlify.app/logistic) - ignore Bayes parts in these. [Hartig DHARMa vignette](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html), [Analyzing % Cover and Porportional Data](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13234)  
__Additional Reading:__ [Maximum Likelihood](https://fw8051statistics4ecologists.netlify.app/mle), [O'Hara and Kotze 2010](http://byrneslab.net/classes/biol607/readings/o'hara_and_kotze_do_not_log_transform.pdf), [Wharton and Hui 2011](http://byrneslab.net/classes/biol607/readings/Wharton_Hui_2011_Ecology.pdf), [Ver Hoef and Boveng 2007 on NB v. Quasi](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/07-0043.1),  [Intro to Beta Regression for % Cover Data](https://rpubs.com/nicoleknight/936037), [Zero-One Inflated Beta Regression](https://journal.r-project.org/archive/2015/RJ-2015-019/index.html), [Feiberg on GLS](https://fw8051statistics4ecologists.netlify.app/gls.html)      
__Etherpad:__ https://etherpad.wikimedia.org/p/607-glm-2023  
__Homework:__ [Surviving the Titanic](homework/generalized_linear_model.html)  

### Week 9   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ [Random Effects](https://biol609.github.io/lectures/07_random_effects.html#/section), [Mixed Models](lectures/mixed_models.html)   
__Data:__ TBA  
__Lab Topic:__ Mixed Models  
__Packages for the Week:__ `install.packages(c("glmmTMB", "lme4", "merTools", "broom.mixed))`  
__Reading:__ [Linear Mixed Effects Models](https://fw8051statistics4ecologists.netlify.app/mixed), [Generalized Linear Mixed Effects Models](https://fw8051statistics4ecologists.netlify.app/glmm)   
__Additional Reading:__ [Generalized linear mixed models: a practical guide for ecology and evolution](https://www.cell.com/trends/ecology-evolution/fulltext/S0169-5347(09)00019-6), [Gelman and Hill Ch 12](https://byrneslab.net/classes/biol609/gelman_hill_ch_12.pdf) login/pass biol609, [Zuur ch. 9](https://byrneslab.net/classes/biol609/Zuur_ch_5.pdf). 
__Your One Stop FAQ:__ [Your One Stop FAQ: Ben Bolker’s Mixed Model’s FAQ](https://bbolker.github.io/mixedmodels-misc/)   
__Etherpad:__ https://etherpad.wikimedia.org/p/607-mixed-2023  

__MIDTERM__: Due Tues Nov 22nd, 5pm. [Get it here](exams/exam_2022.html)  

## Block 3. Experimental and Observational Study Design and Causal Inference

### Week 10.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__  [Structural Causal Models](lectures/causal_diagrams.pptx). [Causal Inference and Experiments](./lectures/causality_experiments.pptx)  
__Reading:__  [Arif et al. on Structural Causal Models](./readings/Arif_et_al_2022.zip), [Kimmel et al. on Experimental Design](./readings/Kimmel_et_al_2021.pdf)  
__Optional Reading:__ [Grace and Irvine on designing SCMs](Grace_and_Irvine_2019.pdf), [A Beastiary of Experimental Designs](readings/Gotelli&EllisonChapter7.pdf)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-causal-2023  
  
  
### Week 11.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Paper Discussion:__  Discussion of [papers](readings/exp_design_papers.zip) - see [here](https://docs.google.com/spreadsheets/d/1r0VfnAJ3tuJufnLnrVC6Nx0qiyP4l6QILarKTsduov0/edit?usp=sharing) for assignments  
__Lecture:__  [Using causal models to get valid inference from observational and quasi-experimental studies](lectures/causal_models_regression.pptx)  
__Reading:__ [The Causal Dag and Haunted Terror](./readings/McElreath_Ch_6.pdf)  
__Optional Reading:__ [Collider Bias and Covid](readings/griffith_collider_bias_covid.pdf)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-obs-2023  
__Lab:__ Exam work time  
  
  
### Week 12.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Paper Discussion:__  Discussion of [papers](readings/obs_sampling.zip) - see [here](https://docs.google.com/spreadsheets/d/1r0VfnAJ3tuJufnLnrVC6Nx0qiyP4l6QILarKTsduov0/edit?usp=sharing) for assignments  
__Homework:__ THANKSGIVING BREAK! EAT PIE!  

## Block 4. Drawing Inference from Studies

  
### Week 13.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Ways of Knowing: [NHT](./lectures/linear_regression_nht), [NHT with Likelihood](lectures/mle_cv.html), [Power Analysis](./lectures/power_analysis.html)    
__Lab Topic:__ [Null Hypothesis Testing](lab/nht.html), [Power Analysis](lab/power_analysis_lab.html)  
__Data for Lab__: [mole rats](lab/data/18e4MoleRatLayabouts.csv), [intertidal algae](lab/data/18e3IntertidalAlgae.csv), [cryptosporidium infection](lab/data/cryptoDATA.csv), [Keeley et al. data](lab/data/Keeley_rawdata_select4.csv)   
__Reading:__ W&S 20, [Muff et al. 2022 on p-values](./readings/muff_et_al_2022.pdf), [Feiberg on Maximum Likelihood](https://fw8051statistics4ecologists.netlify.app/mle.html)     
__Etherpad:__ https://etherpad.wikimedia.org/p/607-nht-eval-2023  
__Packages for The Week:__ `install.packages(c("car", "MASS", "profileModel"))`  


### Week 14-15.   
```{r next_date, echo=FALSE}
```
`r datestring`   

```{r next_date, echo=FALSE}
```
`r datestring`   

__Lecture:__ Ways of Knowing: [Cross-Validation and AIC](./lectures/crossvalidation.html), [Bayesian Data Analaysis](lectures/bayes.html), [Fitting a line with Bayesian techniques](lectures/16_bayesian_lm.html), [Bayesian Inference on a Line](lectures/bayesian_lm_inference.html)  
__Lab Topic:__  [Cross-Validation](lab/crossvalidation.html), [Bayes](lab/08_bayes.html)   
__Reading:__ [AIC in Behavioral Ecology](https://link.springer.com/article/10.1007/s00265-010-1029-6), [Ellison 1996](http://byrneslab.net/classes/biol607/readings/Ellison_1996_ecol_app.pdf), Statistical Rethinking [Chapter 1](http://byrneslab.net/classes/biol607/readings/rethinking_ch_1.pdf) and [Chapter 2](http://byrneslab.net/classes/biol607/readings/rethinking_ch_2.pdf), [Aho et al. 2014](./readings/Aho_2014_ecolog_bic.pdf)    
__Additional Reading__: [Intro to Bayes](https://www.bayesrulesbook.com/chapter-1.html), [Bayes' Rule](https://www.bayesrulesbook.com/chapter-2.html)[how to choose a prior](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations),  [regression models with brms](https://magesblog.com/post/2015-09-01-bayesian-regression-models-using-stan/), [rethinking with brms](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/) (many very cool examples), [brms tutorials](https://www.rensvandeschoot.com/tutorials/brms/),  [Bayesian basics with R](https://m-clark.github.io/bayesian-basics/)  
__Optional Books__ [Model Selection and Multimodel Inference](https://ebookcentral.proquest.com/lib/umboston/reader.action?docID=3035464), [Bayes Rules!](https://www.bayesrulesbook.com/)  
__Packages for The Week:__ `install.packages(c("AICcmodavg", "brms", "tidybayes", ggdist", "bayesplot")`   
__Etherpad:__ https://etherpad.wikimedia.org/p/607-cv-bayes-2023  
