---  
  title: "Course Schedule and Readings"  
---  

<!--


NEXT TIME I TEACH 607 in 2024
-->

  While the topics covered are broad, each week will feature different examples from genetics, ecology, molecular, and evolutionary biology highlighting uses of each individual set of techniques.    
  
  F = Fieberg, W&G = [Wickham and Grolemund](http://r4ds.had.co.nz), U/P for linked pdfs = biol607  
  
  Change .html to .Rmd in your browser to get the markdown of all lectures, etc.

**Recorded Lecture**: All lectures and labs will be recorded and are available at https://www.youtube.com/playlist?list=PLZRMqMK8aRmIMuiEX-QVxLNk2o6SlRa-m   
    
**Turning in Homework**: All homework should be completed using RMarkdown or Quarto. You'll freely mix answers in text and code there. Use projects. Please zip up the full project folder for each to submit so that we have the .rmd (or .qmd), .html output, data, etc. and can recompile your homework if we need to. An ideal project structure would be something like this:
  
homework_X/  
&nbsp; |--- markdown  
&nbsp; |--- scripts  
&nbsp; |--- data  

directory structure, so that all data is in `data/`. **Please zip up the archive for the homework and standardize filenames as follows: `number_lastName_firstName_2023.zip`** where number is the week number the homework is due (from weeks below) and make sure to include the 0s for numbers like 01), and your last and first names - well, you should know them!  
  
To submit homework, use the dropbox link you were sent by the TA.  


```{r dates, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
current_date = parse_date_time("2023-09-04", orders="ymd")
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```    
     
## Block 1: Introduction to Programming and Reproducibility

### Week 1. 
`r datestring`   
__Lecture:__ [Class intro](lectures/01_intro_x.html), [Intro to R](lectures/02_r_intro.html). \
__Lab:__ [Matrices, Lists and Data Frames](lectures/02a_r_intro_list_matrix_df.html). [Introduction to Markdown](https://rmarkdown.rstudio.com/lesson-1.html)    
__Reading:__ W&G   [Intro](https://r4ds.hadley.nz/intro), [Workflow basics](https://r4ds.hadley.nz/workflow-basics), [Scripts and Projects](https://r4ds.hadley.nz/workflow-scripts), [Code Style](https://r4ds.hadley.nz/workflow-style), [Vectors](http://r4ds.had.co.nz/vectors.html), and [Quarto](https://r4ds.hadley.nz/quarto)   
__Further Exploration:__ [Quarto Formats](https://r4ds.hadley.nz/quarto-formats), https://quarto.org/ for everything!  
__Cheat Sheets:__ [Quarto Cheat Sheet](https://rstudio.github.io/cheatsheets/quarto.pdf)  
__Install R:__ Go to https://cloud.r-project.org/ and get the right version of R for you. Then, go to https://www.rstudio.com/products/rstudio/download/#download and install Rstudio.  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-intro-2023  
__Homework:__ [Intro to R and Data Frames](./homework/intro_to_r.html)  

### Week 2.   
```{r next_date, echo=FALSE}
current_date <- current_date+ days(7)
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```
`r datestring`   
__Lecture:__  [Data visualization](lectures/05_data_viz_principles.pptx)  
__Lab Topic:__ [Introduction to ggplot2](lab/04_ggplot_palmer_intro.html).  
__Reading:__  [Unwin 2008](http://byrneslab.net/classes/biol607/readings/Unwin_2008_dataviz.pdf), W&G Chapters on [Data Vizualization](http://r4ds.had.co.nz/data-visualisation.html) and [Graphics for Communication](http://r4ds.had.co.nz/graphics-for-communication.html),  [Introducing Palmer Penguins](https://journal.r-project.org/articles/RJ-2022-020/)    
__Optional Reading:__  [Friendly 2008 on History of Data Viz](http://byrneslab.net/classes/biol607/readings/Friendly_2008_dataviz_history.pdf), [DC Starting with Data](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html), [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/) - note, this is a whole book, but scan it - it's amazing\
__Etherpad:__ https://etherpad.wikimedia.org/p/dataviz-2023  
__Packages used this Week:__ [ggplot2](https://ggplot2.tidyverse.org/), [ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html), [ggdist](https://mjskay.github.io/ggdist/) - `install.packages(c("ggplot2", "ggridges", "palmerpenguins", "ggdist"))`  
__Cheat Sheets:__  [Ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf), [Choosing a good chart cheat sheet](https://biol355.github.io/Readings/choosing-a-good-chart-09.pdf)  
<!-- __In Class Code:__  [Intro to data and  ggplot2](in_class_code/2020/scripts/ggplot2_intro.R)\ -->
__Homework:__ [ggplot2 homework](./homework/ggplot.html)  
  
### Week 3.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__  [Data import, using libraries](lectures/06_read_data_libraries.html),  [Data manipulation, summarization, and making it tidy](https://datacarpentry.org/R-ecology-lesson/03-dplyr.html)    
<!--, [Joins](lectures/18_join.html)  -->
__Data:__ [Portal Data](https://ndownloader.figshare.com/files/2292169) - and learn more [here](https://portal.weecology.org/)  
__Reading:__  [Data organization in spreadsheets](./readings/Browman_and_Woo_Spreadsheets.pdf), [10 Commandments for Good Data Managament](https://dynamicecology.wordpress.com/2016/08/22/ten-commandments-for-good-data-management/), W&G Chapters on [data import](https://r4ds.hadley.nz/data-import.html),  [pipes](https://r4ds.hadley.nz/workflow-pipes.html), [data transformation](https://r4ds.hadley.nz/data-transform.html), [tidy data](https://r4ds.hadley.nz/data-tidy.html)    
__Optional Reading:__ [Managing Data Frames with the Dplyr package](https://bookdown.org/rdpeng/exdata/managing-data-frames-with-the-dplyr-package.html), [Strings](http://r4ds.had.co.nz/strings.html), [factors](https://r4ds.hadley.nz/factors.html), and [Dates](http://r4ds.had.co.nz/dates-and-times.html)   
__Cheat Sheets:__ [Reading data into R](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf), [Dplyr cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf).  
__Packages:__ `install.packages(c("dplyr", "janitor", "skimr", "lubridate", "tidyr", "readr", "readxl", "tibble"))` - [readr](https://readr.tidyverse.org/),  [readxl](https://readxl.tidyverse.org/),  [tibble](https://tibble.tidyverse.org/),  [skimr](https://docs.ropensci.org/skimr/), 
[janitor](https://garthtarr.github.io/meatR/janitor.html),
[visdat](https://docs.ropensci.org/visdat/)  
__Etherpad:__ http://etherpad.wikimedia.org/p/607-tidy-2023  
__Homework:__ [Data Manipulation and Pivot homework](./homework/dplyr_tidyr.html)  


### Week 4.   
```{r next_date, echo=FALSE }
```
`r datestring`   
__Lecture:__ Descriptive statistics, probability distributions, simulation and sample size - [Functions](lectures/functions.html), [What is a Sample?](lectures/03_sampling_lecture_x.html) and [Precision and Simulation for Estimation](lectures/04_simulation_estimation_x.html).   
__Lab Topic:__ [Sampling and simulation](lab/02_sim_samp.html).    
__Reading:__ W&S 1,3-4,  [functions](https://r4ds.had.co.nz/functions.html), [Cumming et al. 2007 on SDs, SEs, and CIs](http://byrneslab.net/classes/biol607/readings/Cumming_2007_error.pdf)   
__Etherpad:__ https://etherpad.wikimedia.org/p/sampling-2022  
__Packages for the Week:__ [dplyr](https://dplyr.tidyverse.org/) - `install.packages(c("dplyr", "purrr"))`  
__In Class Code:__ [Code from Lab for dplyr](in_class_code/2020/scripts/intro_to_tidyverse.R), [Code from Lab for simulation from a population](in_class_code/2020/scripts/simulation.R), [Code from Lab for Bootstrapped SE](in_class_code/2020/scripts/boot_se.R)\
__Homework:__ [Functions, Iteration, and Precision](./homework/function_iteration.html)  


## Block 2: Linear Models for Data Analysis

### Week 5.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Introduction to Regression: [Correlation and Regression](lectures/linear_regression.html), [Fit and Precision](lectures/linear_regression_details.html)  
__Lab Topic:__ [Linear regression, diagnostics, visualization](lab/06_lm.html), and [data](lab/data_06.zip) \
__Reading:__ [Fieberg linear regression review](https://fw8051statistics4ecologists.netlify.app/linreg.html), W&G on [model basics](http://r4ds.had.co.nz/model-basics.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-lm-2022  
<!--__In Class Code:__ [reproducible examples](in_class_code/2020/scripts/reprex.R), [lm](in_class_code/2020/scripts/linear_regression.R)  -->
__Homework:__ [Correlation and Linear Models](homework/06_correlation_regression.html)  


### Week 6   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Categorical Predictors [It
s All Linear Models; t-test edition](lectures/t_as_lm.html),  [Many Levels of a Categorical Predictor](lectures/anova_1.html)      
__Lab Topic:__ [Models with Categorical Variables](lab/categorical_models.html), Final Proposal Meetings	  
__Lab Data:__  [Multiple Files](lab/data/categorical_data.zip)   
__Reading:__ [Common statistical tests are linear models](https://lindeloev.github.io/tests-as-linear/), [Feiberg 3.5](https://fw8051statistics4ecologists.netlify.app/matrixreg#categorical-predictors), W&S Chapters 11, 12, 14      
__Packages for The Week:__ `install.packages(c("car", "emmeans", "multcompView", "contrast"))`    
__Etherpad:__ https://etherpad.wikimedia.org/p/607-categorical_lm-2022  
__Homework:__ [Models with Categorical Variables](./homework/categorical.html)  

<!-- __In Class Code:__ [comparing two means](./in_class_code/2020/scripts/comparing_means.R)  -->  


### Week 7   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lectures:__ May Predictors: [Many Predictors](./lectures/many_predictors.html), [Many categories or mixing categories and continuous predictors](./lectures/many_types_of_predictors.html)   
__Lab Topic:__  [Models with many types of predictors](lab/many_predictor_types.html)  
__Lab Data:__  [Multiple Files](lab/data_10.zip)  
__Reading:__ [Feiberg ch 3](https://fw8051statistics4ecologists.netlify.app/matrixreg), W&S 15, 18   
__Optional Reading:__ [Analysis of variance with unbalanced data: an update for ecology & evolution](./readings/Hector_et_al-2010-Journal_of_Animal_Ecology.pdf)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-many-predictors-2022  
__Homework:__ [Many Predictors!](homework/many_predictors.html), [RIKZ Data](homework/data/rikz.csv)  
__Project Proposal:__ [general instructions here](project_proposal.html)  


### Week 8.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lectures:__ [Interaction Effects with Categorical Variables](lectures/categorical_interactions.html), [Interaction Effects and Nonlinearities in Any Model](./lectures/23_general_linear_model.html)    
__Lab Topic:__ [Models with Interaction Effects](lab/interacting_predictors.html) - [data files](lab/data_11.zip)  
__Readings:__ Re-Read [Feiberg ch 3](https://fw8051statistics4ecologists.netlify.app/matrixreg), [Feiberg ch 6](https://fw8051statistics4ecologists.netlify.app/multicol.html),  [Simple means to improve the interpretability of regression coefficients](./readings/Schielzeth_2010_MEE.pdf)  
__Optional Readings:__  [Centring in regression analyses: a strategy to prevent errors in statistical inference](./readings/kramer_blasey_centering.pdf)   
__Etherpad:__ https://etherpad.wikimedia.org/p/607-interactions-2022  
__Homework:__ [Interactions](homework/interactions.html)  

### Week 9   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Beyond Linear Regression, But Still Linear [Modeling Error](./lectures/modeling_variance.html), [Generalized Linear Models](lectures/generalized_linear_models.html)   
__Data:__ [GLS Data](./lectures/data/gls/gls_data.zip), [GLM Data](./lab/gzlm_data.zip)  
__Lab Topic:__ [Generalized Linear Models](./lab/12_gzlm.html)  
__Packages for the Week:__ `install.packages(c("MASS", "readxl", "betareg", "DHARMa", "TMB", "glmmTMB"))`  
__Reading:__ [Feiberg on GLS](https://fw8051statistics4ecologists.netlify.app/gls.html), [Feiberg on Generalized Linear Models](https://fw8051statistics4ecologists.netlify.app/glm.html), [Models for Count Data](https://fw8051statistics4ecologists.netlify.app/poissonr), [Feiberg on Logistic Regression](https://fw8051statistics4ecologists.netlify.app/logistic) - ignore Bayes parts in these. [Hartig DHARMa vignette](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html), [Analyzing % Cover and Porportional Data](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13234)  
__Additional Reading:__ [O'Hara and Kotze 2010](http://byrneslab.net/classes/biol607/readings/o'hara_and_kotze_do_not_log_transform.pdf), [Wharton and Hui 2011](http://byrneslab.net/classes/biol607/readings/Wharton_Hui_2011_Ecology.pdf), [Intro to Beta Regression for % Cover Data](https://rpubs.com/nicoleknight/936037), [Zero-One Inflated Beta Regression](https://journal.r-project.org/archive/2015/RJ-2015-019/index.html)    
__Etherpad:__ https://etherpad.wikimedia.org/p/607-glm-2022  

__MIDTERM__: Due Tues Nov 22nd, 5pm. [Get it here](exams/exam_2022.html)  

## Block 3. Experimental and Observational Study Design and Causal Inference

### Week 10.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__  [Structural Causal Models](lectures/causal_diagrams.pptx). [Causal Inference and Experiments](./lectures/causality_experiments.pptx)  
__Reading:__  [Arif et al. on Structural Causal Models](./readings/Arif_et_al_2022.zip), [Kimmel et al. on Experimental Design](./readings/Kimmel_et_al_2021.pdf)  
__Optional Reading:__ [Grace and Irvine on designing SCMs](Grace_and_Irvine_2019.pdf), [A Beastiary of Experimental Designs](readings/Gotelli&EllisonChapter7.pdf)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-causal-2022  
  
  
### Week 11.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Paper Discussion:__  Discussion of [papers](readings/exp_design_papers.zip) - see [here](https://docs.google.com/spreadsheets/d/1r0VfnAJ3tuJufnLnrVC6Nx0qiyP4l6QILarKTsduov0/edit?usp=sharing) for assignments  
__Lecture:__  [Using causal models to get valid inference from observational and quasi-experimental studies](lectures/causal_models_regression.pptx)  
__Reading:__ [The Causal Dag and Haunted Terror](./readings/McElreath_Ch_6.pdf)  
__Optional Reading:__ [Collider Bias and Covid](readings/griffith_collider_bias_covid.pdf)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-obs-2022  
__Lab:__ Exam work time  
  
  
### Week 12.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Paper Discussion:__  Discussion of [papers](readings/obs_sampling.zip) - see [here](https://docs.google.com/spreadsheets/d/1r0VfnAJ3tuJufnLnrVC6Nx0qiyP4l6QILarKTsduov0/edit?usp=sharing) for assignments  
__Homework:__ THANKSGIVING BREAK! EAT PIE!  

## Block 4. Drawing Inference from Studies

  
### Week 13.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Ways of Knowing: [NHT](./lectures/linear_regression_nht), [NHT with Likelihood](lectures/mle_cv.html), [Power Analysis](./lectures/power_analysis.html)    
__Lab Topic:__ [Null Hypothesis Testing](lab/nht.html), [Power Analysis](lab/power_analysis_lab.html)  
__Data for Lab__: [mole rats](lab/data/18e4MoleRatLayabouts.csv), [intertidal algae](lab/data/18e3IntertidalAlgae.csv), [cryptosporidium infection](lab/data/cryptoDATA.csv), [Keeley et al. data](lab/data/Keeley_rawdata_select4.csv)   
__Reading:__ W&S 20, [Muff et al. 2022 on p-values](./readings/muff_et_al_2022.pdf), [Feiberg on Maximum Likelihood](https://fw8051statistics4ecologists.netlify.app/mle.html)     
__Etherpad:__ https://etherpad.wikimedia.org/p/607-nht-eval-2022  
__Packages for The Week:__ `install.packages(c("car", "MASS", "profileModel"))`  


### Week 14-15.   
```{r next_date, echo=FALSE}
```
`r datestring`   

```{r next_date, echo=FALSE}
```
`r datestring`   

__Lecture:__ Ways of Knowing: [Cross-Validation and AIC](./lectures/crossvalidation.html), [Bayesian Data Analaysis](lectures/bayes.html), [Fitting a line with Bayesian techniques](lectures/16_bayesian_lm.html), [Bayesian Inference on a Line](lectures/bayesian_lm_inference.html)  
__Lab Topic:__  [Cross-Validation](lab/crossvalidation.html), [Bayes](lab/08_bayes.html)   
__Reading:__ [AIC in Behavioral Ecology](https://link.springer.com/article/10.1007/s00265-010-1029-6), [Ellison 1996](http://byrneslab.net/classes/biol607/readings/Ellison_1996_ecol_app.pdf), Statistical Rethinking [Chapter 1](http://byrneslab.net/classes/biol607/readings/rethinking_ch_1.pdf) and [Chapter 2](http://byrneslab.net/classes/biol607/readings/rethinking_ch_2.pdf), [Aho et al. 2014](./readings/Aho_2014_ecolog_bic.pdf)    
__Additional Reading__: [Intro to Bayes](https://www.bayesrulesbook.com/chapter-1.html), [Bayes' Rule](https://www.bayesrulesbook.com/chapter-2.html)[how to choose a prior](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations),  [regression models with brms](https://magesblog.com/post/2015-09-01-bayesian-regression-models-using-stan/), [rethinking with brms](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/) (many very cool examples), [brms tutorials](https://www.rensvandeschoot.com/tutorials/brms/),  [Bayesian basics with R](https://m-clark.github.io/bayesian-basics/)  
__Optional Books__ [Model Selection and Multimodel Inference](https://ebookcentral.proquest.com/lib/umboston/reader.action?docID=3035464), [Bayes Rules!](https://www.bayesrulesbook.com/)  
__Packages for The Week:__ `install.packages(c("AICcmodavg", "brms", "tidybayes", ggdist", "bayesplot")`   
__Etherpad:__ https://etherpad.wikimedia.org/p/607-cv-bayes-2022  
