---  
  title: "Course Schedule and Readings"  
---  
<!--


NEXT TIME I TEACH 607
1. Intro to R
2. Plotting
3. Functions, tidy data, dplyr
4. Sampling and Simulation

-->
  While the topics covered are broad, each week will feature different examples from genetics, ecology, molecular, and evolutionary biology highlighting uses of each individual set of techniques.    
  
  W&S = Whitlock and Schluter, W&G = [Wickham and Grolemund](http://r4ds.had.co.nz), U/P for linked pdfs = biol607  
  
  Change .html to .Rmd in your browser to get the markdown
  
**Turning in Homework**: All homework should be completed using RMarkdown. You'll freely mix answers in text and code there. Please submit both the .Rmd and .html output of your homework. If there are data files associated with your homework, when working on it, please make sure you are using the   
  
homework/  
&nbsp; |--- markdown  
&nbsp; |--- data  

directory structure, so that all data is in `../data/` relative to where your homework markdown and outputs are. Please standardize filenames as follows: `number_lastName_firstName_2020.Rmd` where number is the homework number (you'll see it in the homework assignment's filename - and make sure to include the 0s for numbers like 01), and your last and first names - well, you should know them!  
  
To submit homework, go to https://www.dropbox.com/request/MgLy0OxePC4QN8ZDgQ0X and upload the files. 


```{r dates, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
current_date = parse_date_time("2020-09-07", orders="ymd")
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```    
     
## Block 1: Introduction to Programming and Reproducibility

### Week 1. 
`r datestring`   
__Lecture:__ [Class intro](lectures/01_intro_x.html), [Intro to R](lectures/02_r_intro.html). \
__Lab:__ [Matrices, Lists and Data Frames](lectures/02a_r_intro_list_matrix_df.html). [Introduction to Markdown](https://rmarkdown.rstudio.com/lesson-1.html)    
__Reading:__ W&G  [Preface](http://r4ds.had.co.nz/introduction.html), [Intro](http://r4ds.had.co.nz/explore-intro.html), [Workflow basics](http://r4ds.had.co.nz/workflow-basics.html), [Vectors](http://r4ds.had.co.nz/vectors.html), and [Markdown](http://r4ds.had.co.nz/r-markdown.html) Chapters  
__Cheat Sheets:__ [RMarkdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf), [RMarkdown Options Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)    
__In Class Code:__ [Code from Lab](./in_class_code/2020/scripts/my_first_r_script.R), [matrices and lists](./in_class_code/2020/scripts/matrices_lists.R), [markdown intro](./in_class_code/2020/markdown/my_first_markdown.R)\
__Install R:__ Go to https://cloud.r-project.org/ and get the right version of R for you. Then, go to https://www.rstudio.com/products/rstudio/download/#download and install Rstudio.  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-intro-2020  
__Homework:__ [Intro to R and Data Frames](./homework/intro_to_r.html)  

  
### Week 2.   
```{r next_date, echo=FALSE }
current_date <- current_date+ days(7)
datestring <- paste(month(current_date), day(current_date), year(current_date), sep="/")
```
`r datestring`   
__Lecture:__ Descriptive statistics, and the creation of good observational sampling designs.    - [Sampling](lectures/03_sampling_lecture_x.html) and [Simulation for Estimation](lectures/04_simulation_estimation_x.html).   
__Lab Topic:__ [Sampling and simulation](lab/02_sim_samp.html). Libraries in R. Dplyr.   
__Reading:__ W&S 1,3-4, W&G Chapters on [data transformation](http://r4ds.had.co.nz/transform.html) and [pipes](http://r4ds.had.co.nz/pipes.html),  [Dplyr cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdfPaperpile)  
__Optional Reading:__ [Cumming et al. 2007 on SDs, SEs, and CIs](http://byrneslab.net/classes/biol607/readings/Cumming_2007_error.pdf), [Simpler Coding with Pipes](https://www.r-statistics.com/2014/08/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/), [Managing Data Frames with the Dplyr package](https://bookdown.org/rdpeng/exdata/managing-data-frames-with-the-dplyr-package.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/sampling-2020  
__Packages for the Week:__ [dplyr](https://dplyr.tidyverse.org/) - `install.packages(c("dplyr", "purrr"))`  
__In Class Code:__ [Code from Lab for dplyr](in_class_code/2020/scripts/intro_to_tidyverse.R), [Code from Lab for simulation from a population](in_class_code/2020/scripts/simulation.R), [Code from Lab for Bootstrapped SE](in_class_code/2020/scripts/boot_se.R)\
__Homework:__ [R, Dplyr, and Iteration](./homework/02_iteration.html)  


### Week 3.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ More on simulation of sample properties, [Data visualization](lectures/05_data_viz_principles.pptx)  
__Lab Topic:__ [Introduction to ggplot2](lab/04_ggplot_palmer_intro.html).  
__Reading:__ W&S Chapter 2, [Unwin 2008](http://byrneslab.net/classes/biol607/readings/Unwin_2008_dataviz.pdf), W&G Chapters on [Data Vizualization](http://r4ds.had.co.nz/data-visualisation.html) and [Graphics for Communication](http://r4ds.had.co.nz/graphics-for-communication.html), [DC Starting with Data](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html)  
__Optional Reading:__  [Friendly 2008 on History of Data Viz](http://byrneslab.net/classes/biol607/readings/Friendly_2008_dataviz_history.pdf), [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/) - note, this is a whole book, but scan it - it's amazing\
__Etherpad:__ https://etherpad.wikimedia.org/p/dataviz-2020  
__Packages used this Week:__ [ggplot2](https://ggplot2.tidyverse.org/), [ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html), [ggdist](https://mjskay.github.io/ggdist/) - `install.packages(c("ggplot2", "ggridges", "palmerpenguins", "ggdist"))`  
__Cheat Sheets:__  [Ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf), [Choosing a good chart cheat sheet](https://biol355.github.io/Readings/choosing-a-good-chart-09.pdf)  
__In Class Code:__  [Intro to data and  ggplot2](in_class_code/2020/scripts/ggplot2_intro.R)\
__Homework:__ [ggplot2 homework](./homework/03_samp_ggplot.html)  
  
### Week 4.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ [Functions](lectures/functions.html), [Data import, libraries, factors and forcats](lectures/06_read_data_libraries.html),  [Tidy data](lectures/17_tidy.html), [Joins](lectures/18_join.html)  
__Data:__ [Hemlock](lectures/data/18/hemlock.xlsx)  
__Reading:__  [Data organization in spreadsheets](./readings/Browman_and_Woo_Spreadsheets.pdf), [10 Commandments for Good Data Managament](https://dynamicecology.wordpress.com/2016/08/22/ten-commandments-for-good-data-management/), W&G Chapters on [functions](https://r4ds.had.co.nz/functions.html), [tidy data](http://r4ds.had.co.nz/tidy-data.html), [Strings](http://r4ds.had.co.nz/strings.html), and [Dates](http://r4ds.had.co.nz/dates-and-times.html)   
__Cheat Sheets:__ [Reading data into R](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf).  
__Packages:__ `install.packages(c("janitor", "skimr", "lubridate", "tidyr", "readr", "readxl", "tibble"))` - [readr](https://readr.tidyverse.org/),  [readxl](https://readxl.tidyverse.org/),  [tibble](https://tibble.tidyverse.org/),  [skimr](https://docs.ropensci.org/skimr/), 
[janitor](https://garthtarr.github.io/meatR/janitor.html),
[visdat](https://docs.ropensci.org/visdat/)
__Etherpad:__ http://etherpad.wikimedia.org/p/607-tidy-2020  
__In Class Code:__ [functions](in_class_code/2020/scripts/functions.R), [loading data](in_class_code/2020/scripts/loading_data.R), [data reshaping](in_class_code/2020/scripts/tidyr.R), [created data](in_class_code/2020/data/test_data.xlsx)    
__Homework:__ [Functions and Pivot homework](./homework/04_fun_pivot.html)  

## Block 2: Regression and Inference

### Week 5.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Introduction to Regression: [Correlation and Regression](lectures/11_cor_linear_model.html), [Fit and Precision](lectures/12_linear_model_fit.html)  
__Lab Topic:__ [Linear regression, diagnostics, visualization](lab/06_lm.html), and [data](lab/data_06.zip) \
__Reading:__ W&S 16-17, W&G on [model basics](http://r4ds.had.co.nz/model-basics.html), [model building](http://r4ds.had.co.nz/model-building.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-lm-2020  
__In Class Code:__ [lm](in_class_code_2020/scripts/lm_code.R)  
__Homework:__ [Correlation and Linear Models](homework_2020/06_correlation_regression.html)  
  
### Week 6.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Ways of Knowing: NHST and [Likelihood](lectures/13_likelihood.html), [Fitting a line with Likelihood](lectures/14_likelihood_2.html)  
__Lab Topic:__ [Likelihood and linear models](lab/07_likelihood.html), [data](./lab/data/17e8ShrinkingSeals Trites 1996.csv)\
__Reading:__ W&S 20, W&G Chapter [Iteration](http://r4ds.had.co.nz/iteration.html)  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-likelihood-2020  
__In Class Code:__  [linear models with likelihood](in_class_code_2020/scripts/likelihood.R)  
__Packages for The Week:__ `install.packages(c("MASS", "profileModel", "bbmle"))`  
__Homework:__[pufferfish and likelihood](./homework_2020/07_likelihood.html)  

### Week 7.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Ways of Knowing: [Bayesian Inference](lectures/15_bayesian_basics.html), [Fitting a line with Bayesian techniques](lectures/16_bayesian_lm.html)  
__Lab Topic:__ Bayesian computation in R, [Fitting a line with Bayesian techniques](lab/08_bayes.html)   
__Reading:__ [Ellison 1996](http://byrneslab.net/classes/biol607/readings/Ellison_1996_ecol_app.pdf), Statistical Rethinking [Chapter 1](http://byrneslab.net/classes/biol607/readings/rethinking_ch_1.pdf) and [Chapter 2](http://byrneslab.net/classes/biol607/readings/rethinking_ch_2.pdf), [R Users will Now Inevitably Become Bayesians](https://thinkinator.com/2016/01/12/r-users-will-now-inevitably-become-bayesians/), [bayesplot cheat sheet](https://github.com/roualdes/stan_cheatsheets/blob/master/bayesplot/main.pdf)  
__Additional Reading__: [how to choose a prior](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations), [bayesian t-tests](https://vuorre.netlify.com/post/2017/how-to-compare-two-groups-with-robust-bayesian-estimation-using-r-stan-and-brms/), [regression models with brms](https://magesblog.com/post/2015-09-01-bayesian-regression-models-using-stan/), [rethinking with brms](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/) (many very cool examples), [brms tutorials](https://www.rensvandeschoot.com/tutorials/brms/), [How to use rstanarm](https://cran.r-project.org/web/packages/rstanarm/vignettes/rstanarm.html), [Linear Models in rstanarm](https://cran.r-project.org/web/packages/rstanarm/vignettes/lm.html), [Bayesian basics with R](https://m-clark.github.io/bayesian-basics/)  
__Packages for The Week:__ `install.packages(c("brms", "rstanarm", "arm", "rethinking"))` and `devtools::install_github("mjskay/tidybayes")`  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-bayes-2020  
__In Class Code:__ [Bayesian Data Analysis](in_class_code_2020/scripts/bayes.R)  


### Week 8   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Many Predictors [Multiple Regression and the General Linear Model](./lectures/23_general_linear_model.html), [Information Theoretic Approaches](./lectures/24_mmi.html)  
__Lab Topic:__ [Multiple Regression, Multimodel Inference](lab/11_glm_aic.html) - [data files](lab/data_11.zip)  
__Readings:__ [Symonds and Moussalli 2010](http://byrneslab.net/classes/biol607/readings/Symonds_and_Moussalli_2010_behav_ecol.pdf), [Simple means to improve the interpretability of regression coefficients](./readings/Schielzeth_2010_MEE.pdf), [Experimental strategies to assess the biological ramifications of multiple drivers of global ocean change](./readings/Boyd_2020_Glob_Change_Biol.pdf)  
__Optional Readings:__  [Centring in regression analyses: a strategy to prevent errors in statistical inference](./readings/kramer_blasey_centering.pdf), [Model selection for ecologists: the worldviews of AIC and BIC](./readings/Aho_2014_ecolog_bic.pdf), The whole [Ecology Special Section on P Values](http://byrneslab.net/classes/biol607/readings/ecology_forum_on_p_values.pdf) is incredible reading.  
__Etherpad:__ https://etherpad.wikimedia.org/p/607-mlr-2020  
__Packages for the Week:__ `install.packages(c("AICcmodavg", "MuMIn", "modelr"))`  
__In Class Code:__ [ancova, mlr, and aic](./in_class_code_2020/scripts/mlr_int_aic.R)  

## Block 3: Teasing Apart Cause and Effect with Regression Models and Well Designed Studies

### Week 8   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__  Causal Inference and Asking a Good Question
  
### Week 9.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Experimental design and ANOVA [part 1](lectures/19_expts_anova.html), [part 2](lectures/20_anova_2.html), Cottingham discussion    
__Lab Topic:__ [One-Way ANOVA](lab/10_anova.html), Midterm work session	  
__Lab Data:__  [Multiple Files](lab/data_10.zip)  
__Reading:__ W&S Chapter 14-15,  [Cottingham et al. 2005](http://byrneslab.net/classes/biol607/readings/cottingham_et_al_2005_frontiers_all.pdf)  
__Optional Reading:__ [Ecological applications of multilevel analysis of variance](./readings/Qian_et_al-2007-Ecology.pdf) with [appendices](./readings/Qian_Shen.zip)  
__Packages for The Week:__ ` install.packages(c("car", "emmeans", "multcompView", "contrast"))`
__Etherpad:__ https://etherpad.wikimedia.org/p/607-anova-2020
__In Class Code:__ [lots of anova](./in_class_code_2020/scripts/anova.R)  
  
  
### Week 10.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lectures:__ Experimental Design in a Multicausal World - [Multiway ANOVA](lectures/21_multiway_anova.html), [Factorial ANOVA](lectures/22_factorial_anova.html)    
__Lab Topic:__  Discussion of Hurlbert, Cottingham, [Factorial ANOVA](lab/11_anova.html)  
__Lab Data:__  [Multiple Files](lab/data_10.zip)  
__Reading:__ W&S 18, [Hurlbert 1984](http://byrneslab.net/classes/biol607/readings/Hurlbert_1984_eco_mono.pdf)  
__Optional Reading:__ [Analysis of variance with unbalanced data: an update for ecology & evolution](./readings/Hector_et_al-2010-Journal_of_Animal_Ecology.pdf)  
__In Class Code:__ [lots of anova](./in_class_code_2020/scripts/anova_two.R)  
   
## Block 4: Modeling Error and Further Adventures in Regression Models
   
### Week 11.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Modeling Error: [Generalized Least Squares and Heteroscedasticity](./lectures/gls.html), [Entering a non-normal world](lectures/25_generalized_linear_models.html) - Modeling count data with Genearlized linear models.  
__Data:__ [GLS Data](./lectures/data/gls/gls_data.zip), [GLM Data](./lab/gzlm_data.zip)  
__Lab Topic:__ [Generalized Linear Models](./lab/12_gzlm.html)  
__Packages for the Week:__ `install.packages(c("MASS", "readxl", "betareg", "DHARMa"))`  
__Reading:__ [O'Hara 2009](http://byrneslab.net/classes/biol607/readings/O'Hara_2009_glmm.pdf) through section on GLMs, [O'Hara and Kotze 2010](http://byrneslab.net/classes/biol607/readings/o'hara_and_kotze_do_not_log_transform.pdf), [Wharton and Hui 2011](http://byrneslab.net/classes/biol607/readings/Wharton_Hui_2011_Ecology.pdf), [Hartig DHARMa vignette](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html)  
  
  
### Week 12.   
__Lecture:__ Old Wine in a New Bottle: It's all a line, Classical tests in a Regression Framework


## Block 5: Tying it Together Further Adventures in Computation and Presentation


### Week 14.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ [Iteration with purrr](./lectures/iteration.html), Class's Choice   
__Data:__ [gapminder](./lectures/data/gapminder.zip)  
__Reading:__ [Murtaugh 2007](./readings/Murtaugh_2007_Ecology.pdf), [iteration in R4DS](https://r4ds.had.co.nz/iteration.html), [functionals in adv-r](https://adv-r.hadley.nz/functionals.html)  
__Cheat Sheets:__ [Apply Functions with purrr Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/purrr.pdf)  
__Lab Topic:__ Class's Choice, Final Presentation Open Lab  
__In Class Code__: [spread and gather](./in_class_code_2020/scripts/tidyr.R)  

### Week 14.   
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Shiny!
  
### Week 15.  
```{r next_date, echo=FALSE}
```
`r datestring`   
__Lecture:__ Final Presentations  
