---
title: "Many Predictors Homework"
author: "Biol607"
format: 
  html:
    toc: true

execute:
  echo: false
  message: false
  warning: false
  eval: false
---

## 1. Replicated Regression and Interaction Effects

One of my favorite papers on experimental design is [Cottingham et al. 2005](http://byrneslab.net/classes/biol607/readings/cottingham_et_al_2005_frontiers_all.pdf) which posits that regression can actually be a better way to get inference than many categorical levels of treatments - most of the time. If your treatment effects are nonlinear, categorical predictors are better, as they can accomodate any shape of a curve. They thus advocate for 'replicated regression' - a continuous experimental design with replication at each level of the treatment. (Note: we'll be reading this paper in a few weeks, but, feel free to dig in now if you'd like).

In lab, we looked at an experiment from my time as a graduate student manipulating predators in kelp forest mesocosms - [data here](https://biol607.github.io/lab/data/kelp_pred_div_byrnesetal2006.csv) and, if you want, [paper here](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1461-0248.2005.00842.x). This was, in some ways, a replicated regression design, with `Predator_Diversity` as the continuous treatment (0, 1, or 3 species) and `Porp_Change` as the response.

**1.0** Filter out the `no herbivore` treatment - it has an NA for diversity and is not useful here. Also..... is trial really continuous?

**1.1** Fit the appropriate two models (remember, the experiment was run in three `Trial`s) with diversity as a continuous versus discrete variable (i.e., one model with continuous, one model with discrete). Do the assumption tests differ, or is everything OK? If you are *really* worried, note, you can also just use `Change_g` as the response variable with `Before_g` as a covariate (weight before versus change in weight) - it's what I actually did in the paper.

```{r}
library(readr)
library(dplyr)
library(performance)
library(ggplot2)

kelp <- read_csv("https://biol607.github.io/lab/data/kelp_pred_div_byrnesetal2006.csv") |>
  mutate(Trial = as.character(Trial),
         Predator_Diversity_f = as.factor(Predator_Diversity))

#continuous model
mod_c <- lm(Change_g ~ Trial*Predator_Diversity + Before_g, data = kelp)
check_model(mod_c) # looks good  

mod_f <- lm(Change_g ~ Trial*Predator_Diversity_f + Before_g, data = kelp)
check_model(mod_f) # looks good  

```

**1.2** If you compare results - in both - asking if the diversity effect differed by trial (you'll need both `emmeans` and `emtrends` here) does it really look like there is an interaction with trial? If not, refit the model with no interaction!

```{r}
library(emmeans)
emtrends(mod_c, ~   Trial, var = "Predator_Diversity") |>
  contrast("pairwise") |>
  confint() #nope, no difference


emmeans(mod_f, ~   Predator_Diversity_f | Trial) |>
  contrast("pairwise") |>
  confint() #nope, no difference

#refit and check
mod_c <- lm(Change_g ~ Trial+Predator_Diversity + Before_g, data = kelp)
check_model(mod_c) # looks good  

mod_f <- lm(Change_g ~ Trial+Predator_Diversity_f + Before_g, data = kelp)
check_model(mod_f) # looks good  
```

**1.3** OK, with the two models you now have, what different inferences about the role of predator diversity do they give you, seemingly? Use one or more of the tools you have at your disposal - from things in the `emmeans` package (hey, check out that `at` argument for `emmeans()`...) to visualizations to make a convincing case. 

```{r}
# what is different
emmeans(mod_f, ~   Predator_Diversity_f ) |>
  contrast("pairwise") |>
  confint("none") #1 and 3 species the same?


emmeans(mod_c, ~   Predator_Diversity, 
        at = list(Predator_Diversity = c(0,1,3) )) |>
  contrast("pairwise") |>
  confint() # all are different

#The regression shows a positive relationship, but, the factor shows that maybe 1 and 3 are not different? 

```

**1.4** IMPRESS YOURSELF - if you really want to get into the weeds, in the paper, I also did comparisons between indivity monocultures (1 species treatments, which varied in identity) to the polyculture (the mixture) to ask if the polyculture was better than the best performing monoculture - something in the Biodiversity Ecosystem Function literature known as the Overyielding. Do you find evidence of that. Note - this is ticky, so, impress yourself if you want, but, don't worry if you want to move on.


## 2. Interactions with Continuous Variables

Scientists wanted to simulate how different biological interactions might influence the carbon burial potential of sinking algae in the deep ocean. Let's use [this simulated data](data/c_burial_sims.csv) which features sinking rate, microbial abundance, and detritovore abundance as predictors of net carbon sequestration.

**2.1** Load the data, inspect it, and fit a model with a 3-way interaction, Do you meet assumptions?

**2.2** Now the fun part - inference. What do the coefficients tell you?

**2.3** OK - that's a lot. Use your skills of visualization do tease out what the data is telling us. You can use `visreg()` or `augment()` with `data_grid()` or whatever you would like. Make this model make sense so that you can tell your audience how these three parameters work together to influence carbon burial!



----


## Meta 1. 
What do you find most interesting about intereaction effects? What do you find most intimidating?

## Meta 2. 
Where do you see interaction effects fitting into your own research?

## Meta 3.
Now that we have fully explored purely "linear" models, what one question or concern do you still have?

## Meta 4. 
How much time did this take you, roughly? Again, I'm trying to keep track that these assignments aren't killer, more than anything.  

## Meta 5. 
Please give yourself a weak/sufficient/strong assessment on this assigment. Feel free to comment on why.

