---
title: '%in%'
author: "Biol 607"
date: "10/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Boolean Comparisons of Strings

So, we're used to boolean comparisons of strings. Is something equal to something else?

```{r}
a <- "hello"

a == "goodbye"
```

But what about a vector?


```{r}
a == c("hello", "goodbye")
```

That works, as it does element by element comparison. 

But what about two vector?

```{r}
b <- c("hello", "goodbye")

b == c("hello", "goodbye", "ciao")
```

That.... kind of works. Odd with the false at the end. It's looking at the second, not the first vector. As happened above.

What if we want to know which elements of vector x are in vector y?

The `%in%` operator is great as it asks if elements of one vector are IN the other. `x %in% y`

```{r}
b %in% c("hello", "goodbye", "ciao")
```

Excellent.

```{r}
x <- letters[1:7]
y <- letters[5:15]

x %in% y
```

This works great in a filter, as you can as if a column matches a vector of interest.