---
title: "General Linear Models"
author: "Biol 607"
date: "December 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. ANCOVA
Combining categorical and continuous variables is not that different from multiway ANOVA. To start with, let's look at the neanderthal data.
```{r neand}
neand <- read.csv("./data/11/18q09NeanderthalBrainSize.csv")
head(neand)
```

We can clearly see both the categorical variable we're interested in, and the covariate.

To get a preliminary sense of what's going on here, do some exploratory visualization with `ggplot2` why doncha!

```{r plot_neand}
library(ggplot2)

qplot(lnmass, lnbrain, color=species, data=neand) +
  stat_smooth(method="lm")
```

Now, the CIs are going to be off as this wasn't all tested in the same model, but you begin to get a sense of whether things are parallel or not, and whether this covariate is important.

What other plots might you produce?

As this is a general linear model, good olde `lm()` is still there for us.
```{r neand_lm}
neand_lm <- lm(lnbrain ~ species + lnmass, data=neand)
```

### 1.1 Testing Assumptions of ANCOVA

In addition to the ususal tests, we need to make sure that the slopes really are parallel. We do that by fitting a model with an interaction and testing it (which, well, if there *was* and interaction, might that be interesting).

First, the usual

```{r neand_tests}
par(mfrow=c(2,2))
plot(neand_lm, which=c(1,2,5))
par(mfrow=c(1,1))

#And now look at residuals by group/predictors
library(car)
residualPlots(neand_lm, tests=FALSE)
```

To test the parallel presumption

```{r neand_parallel}
neand_int <- lm(lnbrain ~ species*lnmass, data=neand)

Anova(neand_int)
```

### 1.2 Assessing results

So, first we have our F-tests.

```{r f_neand}
Anova(neand_lm)
```

Both the treatment and covariate matter.

Second, we might want to compare covariate adjusted groups and/or look at covariate adjusted means.

```{r neand_lsmeans}
library(lsmeans)
adj_means <- lsmeans(neand_lm, specs="species")

#adjusted means
adj_means

#comparisons
contrast(adj_means, method="tukey", adjust="none")
```

If you have an interaction, this method is no longer valid. Instead, you'll need to monkey with your posthocs (if you even want to use them - often we don't) to look at tests at different levels of the covariate.

### 1.3 Visualization

Visualization is funny, as you want to make parallel lines and also get the CIs right. Rather than rely on `ggplot2` to do this natively, we need to futz around a bit with generating predictions

```{r neand_predictions}
neand_predict <- predict(neand_lm, interval="confidence")

head(neand_predict)
```

So, here we have fit values, lower confidence interval, and upper confidence intervals. As we have not fed `predict()` any new data, these values line up with our `neand` data frame, so we can cbind it all together, and then use these values to make a prediction plot.

```{r neand_plot_predict}
neand <- cbind(neand, neand_predict)

ggplot(data=neand) +
  geom_point(mapping=aes(x=lnmass, y=lnbrain, color=species)) +
  geom_line(mapping = aes(x = lnmass, y=fit, color=species)) + 
  geom_ribbon(data=neand, aes(x = lnmass, 
                              ymin=lwr, 
                              ymax=upr,
                              group = species), 
              fill="lightgrey", 
              alpha=0.5) +
  theme_bw()
```

And there we have nice parallel lines with model predicted confidence intervals!

### 1.4 Examples
I've provided two data sets:  
1) `18e4MoleRatLayabouts.csv` looking at how caste and mass affect the energy mole rates expend  
2) `18q11ExploitedLarvalFish.csv` looking at how status of a marine area - protected or not - influences the CV around age of maturity of a number of different fish (so, age is a predictor)

Using the following workflow, analyze these data sets.

```{r sampe_workflow}
# Load the data

# Perform a preliminary visualization

# Fit an ANCOVA model

# Test Asssumptions and modeify model if needed

# Evaluate results

# Post-hocs if you can

# Visualize results
```



## 2. Multiple Linear Regression
## 3. Interaction Effects in MLR
## 4. Information Criteria
