---
title: "Hypothesis Testing"
author: "Biol 607"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
<!-- fix heart bpm problem next year - confusing which SD to use -->

### 1) W&S Chapter 6 questions 15, 21, and 29

<!--
15. 
a. There is no difference between pygmy and continental mammoth femur lenghts.
b. There is no difference between drug versus control weight loss rates.
c. There is no difference in the porportion of babies born with cleft palattes between conditions
d. Christmas versus usual music does not alter how much candy shoppers purchase.
e. Male white collared manakins dance equally when females are absent or present.

21.
a. The 60 participant study.
b. The 100 participant study.
c. The 100 participant study.
d. Two-tailed. We do not have an explicit prediction of upregulation versus downregulation.

29.
a. 95
b. 5

-->

### 2) W&S Chapter 7 question 22 - use R to calculate a p-value

<!--
a. 
```{r p}
p <-  6101/9821

sep <- sqrt((p * (1/p))/9821)
```


The 95% CI is `r p - 1.96*sep` $\le$ `r p` $\le$ `r p + 1.96*sep`

b. No. It is below the 95% CI.

-->

### 3) From the Lab: Many SDs and Alphas
Here's the exercise we started in lab. Feel free to look back copiously at the lab handout if you're getting stuck. Remember, for each step, write-out in comments what you want to do, and then follow behing with code.

Now, let's assume an average population-wide resting heart rate of 80 beats per minute with a standard deviation of 6 BPM.

A given drug speeds people's heart rates up on average by 5 BPM. What sample size do we need to achieve a power of 0.8?

#### 3.1) Start up your simulation
Make a simulated data frame to look at the effects of multiple sample sizes: from 1-20, with 500 simulations per sample size, and also multiple SD values, from 3 through 10 (just 3:10, no need for non-integer values). You're going to want `crossing` with your intitial data frame of just sample sizes and a vector of sd values to start. Then generate samples from the appropriate random normal distribution.

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

#make that df!
heart_sims <- crossing(sims = 1:500, n = 1:20, sd = 3:10)
```  

#### 3.2) Z!
OK, now that you've done that, calculate the results from z-tests. Plot p by sample size, using `facet_wrap` for different SD values.

```{r z_tests}
heart_sims <- heart_sims %>%
  rowwise() %>%
  mutate(samp_mean = mean(rnorm(n, mean = 85, sd = sd)),
         z = (samp_mean-80)/(sd/sqrt(n)),
         p = 2*pnorm(abs(z), 0, 1, lower.tail=FALSE)) %>%
  ungroup()

qplot(n,p, data = heart_sims, alpha = I(0.1)) + facet_wrap(~sd)
```

#### 3.3) P and Power
Now plot power for an alpha of 0.05, but use color for different SD values. Include our threshold power of 0.8.

```{r power}
heart_power <- heart_sims %>%
  group_by(n, sd) %>%
  summarize(power = 1-sum(p>0.05)/n()) %>%
  ungroup()

ggplot(heart_power, 
       aes(x = n, y = power, color = factor(sd))) +
  geom_point() + geom_line() +
  geom_hline(yintercept = 0.8)
```

#### 3.4) Many alphas
Last, use `crossing` again to explore changing alphas from 0.01 to 0.1. Plot power curves with different alphas as different colors, and use faceting to look at different SDs. 

```{r heart_power_alpha}

heart_power_alpha <- crossing(heart_sims, alpha = seq(0.01, 0.1, 0.01)) %>%
  group_by(n, sd, alpha) %>%
  summarize(power = 1-sum(p>alpha)/n()) %>%
  ungroup()

ggplot(heart_power_alpha, 
       aes(x = n, y = power, color = factor(alpha))) +
  geom_point() + geom_line() +
  geom_hline(yintercept = 0.8) +
  facet_wrap(~sd)

```

#### 3.5) What does it all mean? What do you learn about how alpha and SD affect power? 
<!-- These analyses show that, while higher sample size = higher power, it decreases as the SD increases. Similarly, with higher alphas, we also lower our alpha, as it becomes harder to reject the null. -->


#### 3.6) How do you think that changing the effect size would affect power? 
You can just answer this without coding out anything. Based on what we've learned so far - what do you think?

<!-- We should increase our power, a the larger effect sizes should overcome the larger SD. It should be easier to see a difference from the null -->